---
title: "Outbreak_Analysis"
author: "Kathryn Fillman"
format: html
editor: visual
---

# HPAI Outbreak Analysis

## Command Hints:

Add new chunk: *Ctrl+Alt+I*\
Render document: *Shift+Ctrl+k*\
run chunk: *Ctrl+shift+enter*\
*If you want something on a newline, make sure to put 2 spaces after it*

## Import data & libraries

```{r Data & Library Import, echo = FALSE}
library(tidyverse)
library(dplyr)
library(lubridate)
library(readxl)

banding <- read.csv("Outbreak_data/Banding_recov_.csv")
FIPS_codes <- read.csv("Outbreak_data/FIPS_Region_Codes.csv")
world_outbreak_data <- read_excel("Outbreak_data/Outbreak_H5_H5N1_2020_Sept2023_World.xlsx", sheet = "outbreaks")
```

## Data Cleaning

```{r Cleaning World Outbreak data}
world_outbreak_data$Outbreak_start_date <- as_date(world_outbreak_data$Outbreak_start_date)
world_outbreak_data$Outbreak_end_date <- as_date(world_outbreak_data$Outbreak_end_date)
world_outbreak_data$outbreak_dates <- interval(world_outbreak_data$Outbreak_start_date, world_outbreak_data$Outbreak_end_date)
```

## Create List of First US Casess

```{r Create list of first US cases}
First_US_Cases <- filter(world_outbreak_data, iso_code=='USA') %>% select(country_unique_code, level1_name, level2_name, Latitude, Longitude, Outbreak_start_date, outbreak_dates, reporting_level, strain_eng, sero_sub_genotype_eng,  disease_eng) %>% arrange(Outbreak_start_date) %>% distinct(level1_name, .keep_all = TRUE)
```

## Create list of connections to/within the US

```{r}
US_connections <- banding %>% filter_all(any_vars(. %in% 'US')) %>% filter(!(FIPS_ADMIN_banding==FIPS_ADMIN_Recov))  %>% left_join(FIPS_codes, by=c('FIPS_ADMIN_banding'='FIPS.Region.Code')) %>% left_join(FIPS_codes, by=c('FIPS_ADMIN_Recov'='FIPS.Region.Code'))
```

## Find Connections to First State

### Create list of connections between state & worldwide locations

```{r}
state_connections <- US_connections %>% filter(State_Province.x=='South Carolina' | State_Province.y=='South Carolina')
```

### Create list of candidate sources

infectious_date is first day of state infection - minimum time before symptoms arise

```{r}
infectious_date <- as_date(as.numeric (First_US_Cases [First_US_Cases$level1_name== 'South Carolina', "Outbreak_start_date"])-3)
```

Filter outbreak data to include only locations with timely outbreaks Then filter again to include only locations that have connections to that state

```{r}
candidate_sources <- world_outbreak_data %>%
  filter((infectious_date %within% outbreak_dates) & (Outbreak_start_date < infectious_date)) %>%
  filter(iso_code %in% state_connections$ISO.3166.1.alpha.3.x | iso_code %in% state_connections$ISO.3166.1.alpha.3.y)
```

Add columns to identify the state for each source & that state's first outbreak

```{r}
state_lat <- First_US_Cases %>% filter(level1_name=='South Carolina') %>% select(Latitude) %>% as.numeric()
state_long <- First_US_Cases %>% filter(level1_name=='South Carolina') %>% select(Longitude) %>% as.numeric()
candidate_sources <- candidate_sources %>% mutate(state='South Carolina', state_first_infection=(infectious_date+3), state_lat=state_lat, state_long=state_long)
```

## Create connection-finding function

```{r}
connection_finder <- function(state) {
  state_connections <- US_connections %>% filter(State_Province.x==state | State_Province.y==state)
  infectious_date <- as_date(as.numeric(First_US_Cases[First_US_Cases$level1_name==state, "Outbreak_start_date"])-3)
  state_sources <- world_outbreak_data %>%
  filter((infectious_date %within% outbreak_dates) & (Outbreak_start_date < infectious_date)) %>%
    filter(iso_code %in% state_connections$ISO.3166.1.alpha.3.x | iso_code %in% state_connections$ISO.3166.1.alpha.3.y)
  state_lat <- First_US_Cases %>% filter(level1_name==state) %>% select(Latitude) %>% as.numeric()
  state_long <- First_US_Cases %>% filter(level1_name==state) %>% select(Longitude) %>% as.numeric()
  state_sources <- state_sources %>% mutate(state=state, state_first_infection=(infectious_date+3), state_lat=state_lat, state_long=state_long)
  candidate_sources <<- rbind(candidate_sources, state_sources)
}
```

## Create list of potential sources for each state

Use connection_finder function

```{r}
for (level1_name in tail(First_US_Cases$level1_name, -1)) {
  connection_finder(level1_name)
}
```

Select subset of meaningful variables

```{r}
candidate_sources <- candidate_sources %>%
  select(state, state_first_infection, state_lat, state_long, iso_code, country, region, reporting_level, strain_eng, sero_sub_genotype_eng,  disease_eng, Latitude, Longitude, Outbreak_start_date, level3_name, level2_name, level1_name)
```

## Create subset of candidate source locations not inlcuing US

```{r}
outside_US_sources <- candidate_sources %>% filter(country!='United States of America')
```

Number of potential source countries (other than US) and how many potential source events they had
```{r}
n_distinct(outside_US_sources$country)
sort(table(outside_US_sources$country))
```

Do the same for level1_names (not including US)

```{r}
n_distinct(outside_US_sources$level1_name)
tail(sort(table(outside_US_sources$level1_name)))
```

## Export Data

### Write to documents

```{r}
write.csv(candidate_sources,"Candidate_Source_List.csv", row.names = FALSE)
write.csv(First_US_Cases,"US_First_Outbreaks.csv", row.names = FALSE)
```
